Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-08-14 grove-paper.tex
Establish continuity in the single-user figures

\tikzset{graphs/graph style/.style={
 tree layout,
 level sep=1cm,
 nodes={draw, rounded corners=1mm},
 edges={draw, sloped, font={\small}}
}}
\tikzset{graphs/root style/.style={
 draw=none,
 as={\textbullet}
}}
%\tikzset{graphs/every edge/.style={}

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:A simple tree with a hole-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   {}
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:A simple tree with a hole-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}

The editor state for each replica in Grove consists of an expression with holes.
For example, Figure~\ref{fig:A simple tree with a hole} shows the simple arithmetic expression \texttt{2 + \hole}, which
has a hole on the right side of the plus operator.

Internally, Grove represents an expression like this as a directed graph with a distinguished root vertex.
Each vertex represents a term and is labeled with a globally unique identifier and a term constructor.
In this figure, vertices~3 and~6 have the constructors \texttt{plus} and \texttt{num(2)}, respectively.
Each term constructor has an associated set of child positions.
The \texttt{plus} constructor has positions for \texttt{left} and \texttt{right} children
and the \texttt{num(n)} constructors have no child positions.
The root vertex has a distinguished identifier and a constructor with a single child position.
For simplicity, we visually represent the root with a dot.

@@

Each edge is labeled with a globally unique identifier (e.g.,~4 and~7 in the figure) and a child position.
An edge indicates that the destination vertex is a child of the origin vertex at the given position.
Visually, we indicate the position of an edge by the location of its origin.
For example, Figure~TODO:1(b) shows the graph corresponding to the term in Figure~TODO:1(a).

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   "$\texttt{var(y)}_5$" [> "right", > "4"']
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, suppose the user adds \texttt{y} as the right child of vertex~1 in Figure~\ref{fig:A simple tree with a hole (Alternate figure)}
so that the expression becomes \texttt{x + y}.
This corresponds to creating a vertex for the variable reference \texttt{y} and adding an edge
to that vertex from vertex~1.
Figure~\ref{fig:Addition} shows the result, where the new vertex has identifier~3.

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   {}
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_3$}]
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  +[as={$\texttt{+}_1$}] -> {
   "$\texttt{var(x)}_3$" [> "left", > "2"'],
   "$\texttt{var(z)}_7$" [> "right", > "6"']
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 "$\texttt{var(y)}_5$" []
};
\end{tikzpicture}
\caption{TODO}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

@@

Next suppose we want to replace \texttt{y} with \texttt{z} so that the code becomes \texttt{x + z}.
This is accomplished by first deleting edge~4.
This results in Figure~\ref{fig:Change-a}, which represents the expression \texttt{x + \hole{}}.
After that deletion, we add a new vertex to represent~\texttt{z} and an
edge from vertex~1 to that vertex.
This results in the bottom graph in Figure~\ref{fig:Change-b},
which represents the expression \texttt{x + z}.

@@

\begin{figure}
\centering
TODO: naive add that creates a conflict
\begin{tikzpicture}
\path (0,0)       node {\includegraphics[scale=0.4]{graphviz/wrapping/1.dot}};
\path (0cm,-4cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/2.dot}};
\path (0cm,-8cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/3.dot}};
\path (0cm,-12cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/4.dot}};
\path (0cm,-16cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/5.dot}};
\end{tikzpicture}
\caption{Wrapping}
\label{fig:Wrapping}
\end{figure}

@@

\begin{figure}
\centering
\begin{tikzpicture}
\path (0,0)    node {\includegraphics[scale=0.75]{graphviz/move/1.dot}};
\path (0,-5cm) node {\includegraphics[scale=0.75]{graphviz/move/2.dot}};
\path (0,-9cm) node {\includegraphics[scale=0.75]{graphviz/move/3.dot}};
\end{tikzpicture}
\caption{Move}
\label{fig:Move}
\end{figure}

>>>>

\tikzset{graphs/graph style/.style={
 tree layout,
 level sep=1cm,
 nodes={draw, rounded corners=1mm},
 edges={draw, sloped, font={\small}}
}}
\tikzset{graphs/root style/.style={
 %draw=none,
 as={\textbullet$_0$}
}}
%\tikzset{graphs/every edge/.style={}

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:A simple tree with a hole-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   {}
  }
 }
};
\end{tikzpicture}
\caption{Graph representation}
\label{fig:A simple tree with a hole-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{A simple tree with a hole}
\label{fig:A simple tree with a hole}
\end{figure}

The editor state for each replica in Grove consists of an expression with holes.
For example, Figure~\ref{fig:A simple tree with a hole} shows the simple arithmetic expression \texttt{x * \hole}, which
has a hole on the right side of the plus operator.

Internally, Grove represents an expression like this as a directed graph with a distinguished root vertex.
Each vertex represents a term and is labeled with a globally unique identifier and a term constructor.
In this figure, vertices~2 and~4 have the constructors \texttt{times} and \texttt{var(x)}, respectively.
Each term constructor has an associated set of child positions.
The \texttt{*} constructor has positions for \texttt{left} and \texttt{right} children
and the \texttt{var} constructors have no child positions but takes an argument naming the particular variable referenced
(e.g., \texttt{var(x)} for the variable \texttt{x}).
The root vertex has a distinguished identifier and a constructor with a single child position.
For simplicity, we visually represent the root with a dot.

@@

Each edge is labeled with a globally unique identifier (e.g.,~1 and~3 in the figure) and a child position.
An edge indicates that the destination vertex is a child of the origin vertex at the given position.
Visually, we indicate the position of an edge by the location of its origin.
For example, Figure~TODO:1(b) shows the graph corresponding to the term in Figure~TODO:1(a).

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
TODO
\caption{User interface}
\label{fig:Addition-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(y)}_6$" [> "right", > "5"']
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Addition-b} after adding vertex~6 and edge~5}
\label{fig:Addition-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Addition}
\label{fig:Addition}
\end{figure}

We model edits to the code as sequences of additions and deletions of edges to this graph.
For example, suppose the user adds \texttt{y} as the right child of vertex~2 in
Figure~\ref{fig:A simple tree with a hole}
so that the expression becomes \texttt{x * y}.
This corresponds to creating a vertex for the variable reference \texttt{y} and adding an edge
to that vertex from vertex~2.
Figure~\ref{fig:Addition} shows the result, where the new vertex has identifier~6.

@@

\begin{figure}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   {}
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 y [as={$\texttt{var(y)}_6$}]
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Addition-b} after deleting edge~5}
\label{fig:Change-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{*}_2$" [> "root", > "1"'] -> {
   "$\texttt{var(x)}_4$" [> "left", > "3"'],
   "$\texttt{var(z)}_8$" [> "right", > "7"']
  }
 }
};
\path (2cm,-2cm) graph[graph style] {
 "$\texttt{var(y)}_6$"
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Change-a} after adding vertex~8 and edge~7}
\label{fig:Change-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Change}
\label{fig:Change}
\end{figure}

@@

Next suppose we want to replace \texttt{y} with \texttt{z} so that the code becomes \texttt{x * z}.
This is accomplished by first deleting edge~5.
This results in Figure~\ref{fig:Change-a}, which represents the expression \texttt{x * \hole{}}.
After that deletion, we add a new vertex to represent~\texttt{z} and an
edge from vertex~2 to that vertex.
This results in the bottom graph in Figure~\ref{fig:Change-b},
which represents the expression \texttt{x * z}.

@@

\begin{figure}
% TODO: naive add that creates a conflict
% \begin{tikzpicture}
% \path (0,0)       node {\includegraphics[scale=0.4]{graphviz/wrapping/1.dot}};
% \path (0cm,-4cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/2.dot}};
% \path (0cm,-8cm)  node {\includegraphics[scale=0.4]{graphviz/wrapping/3.dot}};
% \path (0cm,-12cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/4.dot}};
% \path (0cm,-16cm) node {\includegraphics[scale=0.4]{graphviz/wrapping/5.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style]
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Change-b} after deleting edge~1}
\label{fig:Wrapping-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (-1cm,-3cm) graph[graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-a} after adding vertex~8 and edge~7}
\label{fig:Wrapping-b}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   "$\texttt{*}_2$" [> "left", > "9"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   },
   {}
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-b} after adding edge~9}
\label{fig:Wrapping-c}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Wrapping}
\label{fig:Wrapping}
\end{figure}

TODO: dotted lines for deleted edges

TODO: mention that naive add that creates a multi-child is discussed later

@@

\begin{figure}
% \centering
% \begin{tikzpicture}
% \path (0,0)    node {\includegraphics[scale=0.75]{graphviz/move/1.dot}};
% \path (0,-5cm) node {\includegraphics[scale=0.75]{graphviz/move/2.dot}};
% \path (0,-9cm) node {\includegraphics[scale=0.75]{graphviz/move/3.dot}};
% \end{tikzpicture}
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"']
 }
};
\path (0cm,-3cm) graph [graph style] {
 "$\texttt{*}_2$" -> {
  "$\texttt{var(x)}_4$" [> "left", > "3"'],
  "$\texttt{var(z)}_6$" [> "right", > "5"']
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Wrapping-c} after deleting edge~9}
\label{fig:Move-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.3\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_8$" [> "root", > "7"'] -> {
   {},
   "$\texttt{*}_2$" [> "right", > "11"'] -> {
    "$\texttt{var(x)}_4$" [> "left", > "3"'],
    "$\texttt{var(z)}_6$" [> "right", > "5"']
   }
  }
 }
};
\end{tikzpicture}
\caption{Figure~\ref{fig:Move-a} after adding edge~11}
\label{fig:Move-b}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Moving vertex~2 from the left child to the right child of vertex~8}
\label{fig:Move}
\end{figure}
